---
title: "Reproducible Research Final Assignment"
author: "Sakib Shahriar"
date: "March 8, 2019"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#$\color{blue}{\text{Wild, Wet and Wacky - Which Weather Events Is Causing }}$
#$\color{blue}{\text{The Most Devastating Health And Economic Impact}}$
#$\color{blue}{\text{Across the US?}}$


#Synopsis
Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

#Data Processing

The steps below were used to process the data for final analysis.

  
  

```{r }
#download the data file using the URL
if(!file.exists("stormData.csv.bz2")) {
    download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2",
                  destfile = "stormData.csv.bz2", method = "curl")
}
StmData <- read.csv(file="C:/Users/skb67/Desktop/CourseraDS/Reproducible Research/Final_Assignment/repdata_data_StormData_m.csv", header=TRUE, sep=",")

#Only use the necessary variables from  Stmdata
tidyStmData <- StmData[,c('EVTYPE','FATALITIES','INJURIES', 'PROPDMG', 'PROPDMGEXP', 'CROPDMG', 'CROPDMGEXP')]
```

# Results

#####Your data analysis must address the following questions:

#####1. Across the United States, which types of events (as indicated in the \color{red}{\verb|EVTYPE|}EVTYPE variable) are most harmful with respect to population health?
#####2. Across the United States, which types of events have the greatest economic consequences?

  
  

First, lets analyze the impact on civilian fatalities due to various weather events:

  
  
  

```{r }
library(ggplot2)

# harmful event type that links to fatalities
fatal <- aggregate(FATALITIES ~ EVTYPE, data=tidyStmData, sum)
fatal <- fatal[order(-fatal$FATALITIES), ][1:10, ]
fatal$EVTYPE <- factor(fatal$EVTYPE, levels = fatal$EVTYPE)

g <- ggplot(fatal, aes(x = EVTYPE, y = FATALITIES)) + 
    geom_bar(stat = "identity", fill = "purple") + 
    labs(title="Fatalities due to weather events",
         x ="Event", y = "Number of Fatalities")+
    theme(text = element_text(size=10),axis.text.x = element_text(angle=90, hjust=1))
print(g)
```


**$\color{red}{\text{It is quite evident from the above barplot that out  of the top 10 weather events, tornado affects the population the most in  terms of the number }}$**
**$\color{red}{\text{of civilian fatalities caused, with over 500 fatalities. The next closest, Excessive heat, only causes about 2000 fatalities which is less than half of the former.}}$**

  
  
  

Next, we will analyze the impact on civilian injuries due to the various weather events:

  
  
  

```{r }

# harmful event type that links to injuries
inj <- aggregate(INJURIES ~ EVTYPE, data=tidyStmData, sum)
inj <- inj[order(-inj$INJURIES), ][1:10, ]
inj$EVTYPE <- factor(inj$EVTYPE, levels = inj$EVTYPE)

gg2 <-ggplot(inj, aes(x = EVTYPE, y = INJURIES)) + 
    geom_bar(stat = "identity", fill = "purple") + 
    labs(title="Injuries due to weather events",
         x ="Event", y = "Number of Injuries")+
    theme(text = element_text(size=10),axis.text.x = element_text(angle=90, hjust=1))
print(gg2)
```

**$\color{red}{\text{It is quite evident from the above barplot that out  of the top 10 weather events, tornado affects the population the most in  terms of the number of civilian }}$**
**$\color{red}{\text{injuries caused, with around 90000 injuries The next closest, TSTM Wind, only causes about 6000 injuries which is completely insignificant}}$**
**$\color{red}{\text{compared to the former.}}$**

  
  
  
Finally, lets study the impact of weather events on damages of crops and properties:

  
  
  
```{r }

#harmful event types that link to property and crop damage
damg <- aggregate(PROPDMG + CROPDMG ~ EVTYPE, data=tidyStmData, sum)
names(damg) = c("EVTYPE", "TOTALDAMAGE")
damg <- damg[order(-damg$TOTALDAMAGE), ][1:10, ]
damg$EVTYPE <- factor(damg$EVTYPE, levels = damg$EVTYPE)

gg3<-ggplot(damg, aes(x = EVTYPE, y = TOTALDAMAGE)) + 
    geom_bar(stat = "identity", fill = "purple") + 
    labs(title="Property and Crop damages due to weather events",
         x ="Event", y = "Number of Damages in Dollars")+
    theme(text = element_text(size=10),axis.text.x = element_text(angle=90, hjust=1))

print(gg3)
```

  
  
  
**$\color{red}{\text{It is quite evident from the above barplot that out  of the top 10 weather events, tornado causes the mosst damages to property and crops in terms}}$**
**$\color{red}{\text{of value in dollars. Tornado causes more than 3 million dollars worth of damage. Whereas, the next worse weather event, Flash Floods only causes}}$**
**$\color{red}{\text{about 1.5 million dollars worth of damage, which is half of the former. It should be noted that TSTM Wind, Hail, Flood, Thunderstorm wind,}}$**
**$\color{red}{\text{Lighting all causes good amount of damage to properties and crops.}}$**


  
  
  
  
  
  
  
  
  
   
#### $\color{green}{\text{In conclusion from the analysis, tornado is the most devastating weather event across the United States, causing the}}$

#### $\color{green}{\text{most damage in terms of civilian fatalities and injuries, and also property and crop damages. Other notable weather }}$

#### $\color{green}{\text{events include TSTM Wind, Flashflood. Therefore, the US government and weather experts should take extra}}$

#### $\color{green}{\text{precautionary measures when dealing with such weather experts, especially in the case of Tornado.}}$
